Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-12 paper.Rnw
Draft complete, sending to Mallar for final comments.

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas{} approaches improve segmentation accuracy and
precision over regular atlas-based approaches. These approaches often rely on
a large number of such manually segmented atlases (e.g. 30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} 
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar aggreement to
\multiatlas{} approaches. Thus, our method acts as an accurate \multiatlas{}
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template
library, formed from a subset of target images, via transformations estimated by
nonlinear image registration. The resulting segmentations are then propagated
to each target image and fused using a label fusion method.

We conduct two separate Monte Carlo cross-validation experiments comparing \mb{}
and \multiatlas{} whole hippocampal segmentation using differing atlas and
template library sizes, and registation and label fusion methods. The first
experiment is a 10-fold validation (per parameter setting) over 60 subjects
taken from the Alzheimer's Disease Neuroimaging Database (ADNI), and the second
is a five-fold validation over 81 subjects having had a first episode of psychosis.
In both cases, automated segmentations are compared with manual segmentions
following the \prsnr{}-protocol. Using the best settings found from these
experiments, we segment \Sexpr{length(means.complete$RID)} images of the
\adnidataset{} dataset and compare these with segmentations from existing
automated methods: \fsl{}, \freesurfer{}, \maper{}, and \snt.  Finally, we
conduct a leave-one-out cross-validation (LOOCV) of hippocampal subfield
segmentation using five high-resolution manually segmented atlases
\citep{Winterburn2013}.

\textbf{Results:} 
ADNI-XVAL
In the ADNI cross-validation, using 9 atlases \mb{} achieves a mean Dice's
Similarity Coefficient (DSC) score of \Sexpr{mb.kappa.best} with respect to
manual whole hippocampus segmentations, and also exhibits significantly lower
variability in DSC scores than \multiatlas{} segmentation.  In the younger,
psychosis dataset, \mb{} achieves a mean DSC score of \Sexpr{fep.mb.kappa.best}
and produces volumes which agree with manual segmentation volumes better than
those produced by the \freesurfer and \fsl methods (mean difference in volume:
$80mm^3$, $1600mm^3$, and $800mm^3$, respectively). Across the \adnidataset
dataset,

ADNI-1YR
Subfields

In a LOOCV, \mb{} accurately reproduces the subfields... .

\mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt{})
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.

\textbf{Conclusion:} We demonstrate that \mb{} produces accurate hippocampal
segmentations using only 5 atlases over different hippocampal definitions,
disease populations, and acquisition types, as well as showing that accurate
identification of the hippocampal subfields is possible.
\end{abstract}

@@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}). For
instance, with 9 atlases and using \ants for registration and majority vote
fusion, the mean DSC scores for 1, 5, 9 and 17 templates are \Sexpr{k.1t},
\Sexpr{k.5t}, \Sexpr{k.9t}, \Sexpr{k.17t}, respectively.  A maximum
similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion.

@@

We found a close relationship in total bilateral hippocampal volume between all
methods and the \snt{} semi-automated label volumes (Figure
\ref{fig:ADNI-volumes-plot}). Volumes are correlated with Pearson $r > 0.78$
for all methods across disease categories. Within disease categories (Figure
\ref{fig:ADNI-volumes-boxplot}), \mb{} is consistently well correlated to manual
volumes (Pearson $r > 0.85$), but appears to slightly over-estimate the volume
of the AD hippocampus compared to the \snt{} segmentations.

@@

Computations were performed on the gpc supercomputer at the SciNet HPC
Consortium \cite{Loken2010}. SciNet is funded by: the Canada Foundation for
Innovation under the auspices of Compute Canada; the Government of Ontario;
Ontario Research Fund - Research Excellence; and the University of Toronto.

>>>>

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas{} approaches improve segmentation accuracy and
precision over regular atlas-based approaches. These approaches often rely on
a large number of such manually segmented atlases (e.g. 30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} 
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar aggreement to
\multiatlas{} approaches. Thus, our method acts as an accurate \multiatlas{}
approach when using special, hard-to-define atlases that are laborious to
construct. \\

\textbf{Method:} \mb{} works by propagating atlas segmentations to a template
library, formed from a subset of target images, via transformations estimated by
nonlinear image registration. The resulting segmentations are then propagated
to each target image and fused using a label fusion method.

We conduct two separate Monte Carlo cross-validation experiments comparing \mb{}
and \multiatlas{} whole hippocampal segmentation using differing atlas and
template library sizes, and registation and label fusion methods. The first
experiment is a 10-fold validation (per parameter setting) over 60 subjects
taken from the Alzheimer's Disease Neuroimaging Database (ADNI), and the second
is a five-fold validation over 81 subjects having had a first episode of psychosis.
In both cases, automated segmentations are compared with manual segmentions
following the \prsnr{}-protocol. Using the best settings found from these
experiments, we segment \Sexpr{length(means.complete$RID)} images of the
\adnidataset{} dataset and compare these with segmentations from existing
automated methods: \fsl{}, \freesurfer{}, \maper{}, and \snt.  Finally, we
conduct a leave-one-out cross-validation (LOOCV) of hippocampal subfield
segmentation in standard 3T T1 images, using five high-resolution manually
segmented atlases \citep{Winterburn2013}.

\textbf{Results:} In the ADNI cross-validation, using 9 atlases \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) score of \Sexpr{mb.kappa.best} with
respect to manual whole hippocampus segmentations, and also exhibits
significantly lower variability in DSC scores than \multiatlas{} segmentation.
In the younger, psychosis dataset, \mb{} achieves a mean DSC score of
\Sexpr{fep.mb.kappa.best} and produces volumes which agree with manual
segmentation volumes better than those produced by the \freesurfer{} and \fsl{}
methods (mean difference in volume: $80mm^3$, $1600mm^3$, and $800mm^3$,
respectively). Similarily, in the \adnidataset{} dataset, \mb{} produces
hippocampal segmentations well correlated ($r > 0.85$) with \snt{} semi-automated
reference volumes within disease categories, and shows a conservative bias and a
mean difference in volume of $250 mm^3$ across the entire dataset, compared with
\freesurfer{} and \fsl{} which both overesimate volume differences by $2600
mm^3$ and $2800 mm^3$ on average, respectively. Finally, \mb{} segments the CA1,
CA4/DG and subiculum subfields with an DSC overlap scores of 0.56, 0.65, 0.58,
respectively.

\textbf{Conclusion:} We demonstrate that \mb{} produces accurate whole
hippocampal segmentations using at most only 9 atlases with various hippocampal
definitions, disease populations, and image acquisition types, as well we show
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}). For
instance, with 9 atlases and using \ants{} for registration and majority vote
fusion, the mean DSC scores for 1, 5, 9 and 17 templates are \Sexpr{k.1t},
\Sexpr{k.5t}, \Sexpr{k.9t}, \Sexpr{k.17t}, respectively.  A maximum
similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion.

@@

We found a close relationship in total bilateral hippocampal volume between all
methods and the \snt{} semi-automated label volumes (Figure
\ref{fig:ADNI-volumes-plot}). Volumes are well correlated ($r > 0.78$)c for all
methods, and across disease categories. Within disease categories (Figure
\ref{fig:ADNI-volumes-boxplot}), \mb{} is consistently well correlated to \snt{}
volumes ($r > 0.85$), but appears to slightly over-estimate the volume of the AD
hippocampus compared to the \snt{} segmentations.

@@

Computations were performed on the gpc supercomputer at the SciNet HPC
Consortium \citep{Loken2010}. SciNet is funded by: the Canada Foundation for
Innovation under the auspices of Compute Canada; the Government of Ontario;
Ontario Research Fund - Research Excellence; and the University of Toronto.
